On the instability and degeneracy of deep learning models

### 2019-06-11 writing/paper-note/chapters/instab.Rmd
getting ready for another submission
started edits

In the model framework, let $\boldsymbol X_N = (X_1, \dots, X_N)$ denote a collection of discrete random variables with a finite sample space, $\mathcal{X}^N$, represented as some $N$-fold Cartesian  product. That is, $\mathcal{X}$ with $|\mathcal{X}| < \infty$ denotes the set of potential outcomes for each single variable $X_i$, so that the product space $\mathcal{X}^N$ corresponds to values for the variables $\boldsymbol X_N=(X_1,\ldots,X_N)$. For each $N$, let $P_{\boldsymbol \theta_N}$ denote a probability model on $\mathcal{X}^N$, under which $P_{\boldsymbol \theta_N}(x_1, \dots, x_N) > 0$ is the probability of the data outcome $(x_1, \dots, x_N) \in \mathcal{X}^N$. In this, we assume that the model support of $P_{\boldsymbol \theta_N}$ is the sample space $\mathcal{X}^N$. This framework produces probability models $P_{\boldsymbol \theta_N}$, indexed by a generic sequence of parameters $\boldsymbol \theta_N$, to describe data $\boldsymbol X_N$ of any given sample size $N \geq 1$. For simplicity, we will refer to this distributional class as *Finite Outcome Everywhere Supported (FOES)* models in the following. The dimension and structure of such parameters are generic, without restriction, though natural cases will be seen to include those where $\boldsymbol \theta_N \in \mathbb{R}^{q(N)}$ for some arbitrary integer-valued function $q(\cdot) \geq 1$.

@@

Many model families fall under the umbrella of FOES models. For illustration, this section presents three specific examples of FOES models, including models with deep architectures.

@@

For random variables $\boldsymbol X \equiv\boldsymbol X_N= (X_1, \dots, X_N)$ with sample space $\mathcal{X}^N$, $|\mathcal{X}| < \infty$, consider an exponential family model for $\boldsymbol X$ with probability mass function given by
\begin{equation}
(\#eq:expo)
p_{N, \boldsymbol \theta}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) - \psi(\boldsymbol \theta)\right], \qquad \boldsymbol x \in \mathcal{X}^N,
\end{equation}
depending on parameter vector $\boldsymbol \theta \in \Theta_N \subset \mathbb{R}^{k}$ and natural parameter function $\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L$ with fixed positive integers $k$ and $L$ denoting their dimensions. Above, $\boldsymbol g_N : \mathcal{X}^N \mapsto \mathbb{R}^L$ is a vector of sufficient statistics, while
$$
\psi(\boldsymbol \theta) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) \right], \qquad \boldsymbol \theta \in \Theta_N\equiv \{\boldsymbol \theta \in \mathbb{R}^k : \psi(\boldsymbol \theta) < \infty \},
$$
denotes the normalizing function with parameter space $\Theta_N$. The natural parameter function $\eta (\boldsymbol \theta)$ has a linear form (i.e., $\eta (\boldsymbol \theta)= \bm{A} \boldsymbol \theta$ for a given $L \times k$ matrix $\bm{A}$) in many common model formulations, though may also be nonlinear (e.g., curved exponential families). In the linear case, $\eta (\boldsymbol \theta) = \boldsymbol \theta$ may be generally assumed in the exponential parameterization with a minor modification to the definition of sufficient statistics $\boldsymbol g_N(\boldsymbol x)$.

Such discrete exponential family models are special cases of the FOES models, as seen by  defining $P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \theta_N}(\boldsymbol x)> 0$,  $\boldsymbol x \in \mathcal{X}^N$, based on \@ref(eq:expo) and a parameter sequence $\boldsymbol \theta_N \in \Theta_N \subset \mathbb{R}^k$. For example, if observations $\boldsymbol X = (X_1,\ldots,X_N)$ correspond to $N$ independent and identically distributed Bernoulli random variables, each indicating a binary $0$-$1$ outcome,  the resulting probabilities have exponential form \@ref(eq:expo)  given by

@@

P_{\boldsymbol \theta_N}(\boldsymbol x) \propto
 \exp\left[\boldsymbol \theta_N \sum_{i=1}^N x_i\right], \qquad \boldsymbol x=(x_1,\ldots,x_N) \in\{0,1\}^N, 

@@

 with sufficient statistic $\boldsymbol g_N(\boldsymbol x)\equiv \sum_{i=1}^N  x_i$ and "log odds ratio" parameter $\boldsymbol \theta_N \equiv  \log[ P_{\boldsymbol \theta_N}(X_i=1)/P_{\boldsymbol \theta_N}(X_i=0) ] \in \mathbb{R}$. More generally, supposing $\boldsymbol X =(X_1,\ldots,X_N)$ represent $N$ independent trials, each assuming an outcome $\{1,\ldots,k\}$ among $k$ possibilities (e.g., a die roll), a multinomial distribution is given by

@@

P_{\boldsymbol \theta_N}(\boldsymbol x) \propto  \exp\left[  \boldsymbol \theta_{N}^T g_N(\boldsymbol x)   \right] =
\exp\left[ \sum_{j=1}^k {\theta_{j,N}} \sum_{i=1}^N \mathbb{I}(x_i=j) \right], \qquad \boldsymbol x  \in\{1,\ldots,k\}^N, 

@@

with sufficient statistic $\boldsymbol g_N(\boldsymbol x)$ involving a count $\sum_{i=1}^N \mathbb{I}(x_i=j)$ for each outcome $j \in \{1,\ldots,k\}$, where $\mathbb{I}(\cdot)$ denotes the indicator function, and parameters $\boldsymbol \theta_N=(\theta_{1,N},\ldots,\theta_{k,N})\in\mathbb{R}^k$ defining log-probability ratios $\theta_{i,N}-\theta_{j,N} =\log [P_{\boldsymbol \theta_N}(X_1=i)/P_{\boldsymbol \theta_N}(X_1=j)]$. In addition to such standard models for discrete independent data, exponential models of FOES type commonly arise with dependent spatial data [@besag1974spatial] and network/relational data [@wasserman1994social; @handcock2003assessing]. For a random graph  or network  with, say, $n$ nodes,  consider $N={n \choose 2}$ random edges where the $i$th edge is associated with a pair of nodes $s_i \equiv \{v_i,u_i\}$  and a binary variable $X_i\in\{0,1\}$ indicating   presence/absence of an edge among the node pair $s_i$, $i=1,\ldots,N$. Here the length $N$ of the edge variable sequence $\boldsymbol X = (X_1,\ldots,X_N)$ increases as a function of node number $n$ and corresponding exponential models often incorporate graph topographical features derived from $\boldsymbol X$. As an example, consider a graph model of exponential/FOES form prescribed by

@@

P_{\boldsymbol \theta_N}(\boldsymbol x) \propto
 \exp\left[\sum_{j=1}^3 \theta_{j,N} g_{j,N}(\boldsymbol x)\right], \quad\qquad \boldsymbol x=(x_1,\ldots,x_N)  \in\{0,1\}^N,

@@

involving the numbers of edges, 2-stars and triangles among an outcome $\boldsymbol x$ given by $g_{1,N}(\boldsymbol x)$, $g_{2,N}(\boldsymbol x)$ and $g_{3,N}(\boldsymbol x)$, respectively, along with $k=3$ real parameters $\boldsymbol \theta_N \equiv (\theta_{1,N},\theta_{2,N},\theta_{3,N})$. For this network model \@ref(eq:mod2) in particular, as well as for more general models of form \@ref(eq:expo),  @schweinberger2011instability considered instability in such exponential models with sequences of fixed parameters $\boldsymbol \theta_N = (\theta_1,\ldots,\theta_k)\in\mathbb{R}^k$, $N \geq 1$, of fixed dimension $k$.

For model sequences $P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \theta_N}(\boldsymbol x)$ of the exponential type \@ref(eq:expo), such as those in \@ref(eq:mod1)-\@ref(eq:mod2), note that the dimension $k$ of the parameter $\boldsymbol \theta_N\in\Theta \subset \mathbb{R}^k$ necessarily remains the same for all sample sizes $N \geq 1$ as the form of the natural parameter function $\eta(\cdot)$ in \@ref(eq:expo) and the number of sufficient statistics $\boldsymbol g_{N}(\boldsymbol x)$ do not depend on $N$. Consequently, $\boldsymbol \theta_N$ lies in a parameter space of fixed Euclidean dimension $k$. However, this aspect need not be true for other types of FOES models considered in Sections \@ref(rbm) - \@ref(deep-learning), where instead the numbers of parameters and sufficient statistics commonly increase with the sample size $N$.

@@

A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, with binary variables being most common [cf. @smolensky1986information]. A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with conditional independence within each layer. Let $\boldsymbol X = (X_1,\ldots,X_N)$ denote the $N$ random variables for visibles with support $\mathcal{X}^N$ and $\boldsymbol H = (H_1,\ldots,H_{N_\mathcal{H}})$ denote the $N_\mathcal{H}$ random variables for hiddens with support $\mathcal{X}^{N_\mathcal{H}}$ where $\mathcal{X} = \{-1,1\}$. For parameters $\boldsymbol \theta_N^{\mathcal{H}} \in \mathbb{R}^{N_\mathcal{H}}$, $\boldsymbol \theta_N^{\mathcal{V}}\in \mathbb{R}^N$, and $\boldsymbol \theta_N^{\mathcal{HV}}$ as a real matrix with dimension $N_\mathcal{H} \times N$, the RBM model for $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)$  has the joint probability mass function

@@

\tilde{P}_{\boldsymbol \theta_N} (\tilde{\boldsymbol x}) = \exp\left[ (\boldsymbol \theta_N^{\mathcal{H}})^T \boldsymbol h + \boldsymbol (\boldsymbol \theta_N^{\mathcal{V}})^T \boldsymbol x + \boldsymbol h^T  \boldsymbol\theta_N^{\mathcal{HV}} \boldsymbol x - \psi(\boldsymbol \theta_N)\right], \quad \tilde{\boldsymbol x} = (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_\mathcal{H}}

@@

\psi(\boldsymbol \theta_N) = \log \sum_{\tilde{\boldsymbol x} \in \{\pm 1\}^{N+N_H} } \exp\left[ (\boldsymbol \theta_N^{\mathcal{H}})^T \boldsymbol h + \boldsymbol (\boldsymbol \theta_N^{\mathcal{V}})^T \boldsymbol x + \boldsymbol h^T  \boldsymbol\theta_N^{\mathcal{HV}} \boldsymbol x\right].

@@

Let $\boldsymbol \theta_N = (\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{V}}, \boldsymbol\theta_N^{\mathcal{HV}} ) \in \Theta_N \equiv \mathbb{R}^{q(N)}$, with $q(N) = N + N_\mathcal{H} + N*N_\mathcal{H}$, denote the parameter vector for the RBM, as indexed by the number $N$ of visible random variables (which may differ from the actual lengths of these parameter vectors). The probability mass function for the visible variables $\boldsymbol X = (X_1, \dots, X_N)$ follows from marginalizing the joint specification to yield

@@

P_{\boldsymbol \theta_N} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \{\pm 1\}^{N_{\mathcal{H}}}} \tilde{P}_{\boldsymbol \theta_N} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \{\pm 1\}^{N}\equiv \mathcal{X}^N.

@@

Here the baseline model \@ref(eq:RBM1) for hidden/visible variables is a linear exponential one in sufficient statistics $(\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)$ using $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)$ from   \@ref(eq:RBM1), but the form differs from the previous exponential models in \@ref(eq:expo) in that the lengths of parameters $\boldsymbol \theta_N$ and statistics  $(\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)$ increase to incorporate more visible variables. That is, in contrast to \@ref(eq:expo), the natural parameter function involved in the RBM model \@ref(eq:RBM1), as the identity mapping of the parameters $\boldsymbol \theta_N\in\mathbb{R}^{q(N)}$, naturally grows in  dimension  $q(N)\to \infty$ to accommodate visible variables $X_1, \dots, X_N$ of increasing sample size $N\to \infty$. Additionally, one may further arbitrarily choose the number $N_\mathcal{H}$ of hidden variables $\boldsymbol H$ in the joint RBM model \@ref(eq:RBM1) to define a marginal model \@ref(eq:RBM2) for the $N$ visible variables $\boldsymbol X$, and the number $N_\mathcal{H}$ of hiddens may also potentially increase with $N$. Because $|\mathcal{X}| = 2$ and $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, the RBM  specification \@ref(eq:RBM2) for visibles $\boldsymbol$ corresponds to a FOES model, while the joint distribution \@ref(eq:RBM1) for $(\boldsymbol X, \boldsymbol H)$ is also a FOES model. As this example also indicates, any model formed by marginalizing a base FOES model class, such as the RBM joint specification \@ref(eq:RBM1), is again a FOES model.

@@

Consider two models with "deep architecture" that contain multiple hidden (or latent) layers in addition to a visible layer of data, namely a deep Boltzmann machine [@salakhutdinov2009deep] and a deep belief network @hinton2006fast]. Let $M$ denote the number of hidden layers included in the model and let $N_{(H,1)}, \dots, N_{(H,M)}$ denote the numbers of hidden variables within each hidden layer. Then the random vector $\tilde{\boldsymbol X} = \{H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}, \dots, H^{(M)}_1, \dots, H^{(M)}_{N_{(H,M)}}, \boldsymbol X\}$ collects both the hidden variables
$\{ H_{i}^{(j)} : i=1,\ldots, N_{(H,j)}, j=1,\ldots,M\}$ and visible variables $\boldsymbol X =(X_1,\ldots,X_N)$ in a deep probabilistic model. Each variable outcome will again lie in $\mathcal{X} = \{-1,1\}$.

>>>>

In the model framework, let $\boldsymbol X_N = (X_1, \dots, X_N)$ denote a collection of discrete random variables with a finite sample space, $\mathcal{X}^N$, represented as some $N$-fold Cartesian  product. That is, $\mathcal{X}$ with $|\mathcal{X}| < \infty$ denotes the set of potential outcomes for each single variable $X_i$, so that the product space $\mathcal{X}^N$ corresponds to values for the variables $\boldsymbol X_N=(X_1,\ldots,X_N)$. For each $N$, let $P_{\ak{\thetaN}}$ denote a probability model on $\mathcal{X}^N$, under which $P_{\ak{\thetaN}}(x_1, \dots, x_N) > 0$ is the probability of the data outcome $(x_1, \dots, x_N) \in \mathcal{X}^N$. In this, we assume that the model support of $P_{\ak{\thetaN}}$ is the sample space $\mathcal{X}^N$. This framework produces probability models $P_{\ak{\thetaN}}$, indexed by a \ak{defining} sequence of parameters $\ak{\thetaN}$, to describe data $\boldsymbol X_N$ of any given sample size $N \geq 1$. For simplicity, we will refer to this distributional class as *Finite Outcome Everywhere Supported (FOES)* models in the following. The dimension and structure of such parameters are generic, without restriction, though natural cases will be seen to include those where $\ak{\thetaN} \in \mathbb{R}^{q(N)}$ for some arbitrary integer-valued function $q(\cdot) \geq 1$.

@@

Many model families \ak{lead to} FOES models. \ak{(For clarity, we use ``model family" to refer to generic class or functional form of parametric distributions, while "model" often refers here to a member of the class for an instantiation or selection of parameters. This distinction is not so important for the examples presented in Sections \ref{discrete-exponential-family-models}-\ref{deep-learning}, but becomes more so in Section \ref{instability-results} where model instability depends on both form and a given parameter configuration, rather than just model class form.)} For illustration, this section presents three specific examples of FOES model \ak{classes}, including models with deep architectures.

@@

For random variables $\boldsymbol X \equiv\boldsymbol X_N= (X_1, \dots, X_N)$ with sample space $\mathcal{X}^N$, $|\mathcal{X}| < \infty$, consider an exponential \ak{model family} for $\boldsymbol X$ with probability mass function given by
\begin{equation}
(\#eq:expo)
p_{N, \boldsymbol \theta}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) - \psi(\boldsymbol \theta)\right], \qquad \boldsymbol x \in \mathcal{X}^N,
\end{equation}
depending on parameter vector $\boldsymbol \theta \in \Theta_{\ak{\thetaidx}} \subset \mathbb{R}^{k}$ and natural parameter function $\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L$ with fixed positive integers $k$ and $L$ denoting their dimensions. Above, $\boldsymbol g_N : \mathcal{X}^N \mapsto \mathbb{R}^L$ is a vector of sufficient statistics, while
$$
\psi(\boldsymbol \theta) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) \right], \qquad \boldsymbol \theta \in \Theta_\ak{\thetaidx}\equiv \{\boldsymbol \theta \in \mathbb{R}^k : \psi(\boldsymbol \theta) < \infty \},
$$
denotes the normalizing function with parameter space $\Theta_{\ak{\thetaidx}}$. The natural parameter function $\eta (\boldsymbol \theta)$ has a linear form (i.e., $\eta (\boldsymbol \theta)= \bm{A} \boldsymbol \theta$ for a given $L \times k$ matrix $\bm{A}$) in many common model formulations, though may also be nonlinear (e.g., curved exponential families). In the linear case, $\eta (\boldsymbol \theta) = \boldsymbol \theta$ may be generally assumed in the exponential parameterization with a minor modification to the definition of sufficient statistics $\boldsymbol g_N(\boldsymbol x)$.

Such discrete \ak{exponential models} are special cases of the FOES models, as seen by  defining $P_{\ak{\thetaN}}(\boldsymbol x)\equiv p_{N, \ak{\thetaN}}(\boldsymbol x)> 0$,  $\boldsymbol x \in \mathcal{X}^N$, based on \@ref(eq:expo) and a parameter sequence $\ak{\thetaN} \in \Theta_{\ak{\thetaidx}} \subset \mathbb{R}^k$. For example, if observations $\boldsymbol X = (X_1,\ldots,X_N)$ correspond to $N$ independent and identically distributed Bernoulli random variables, each indicating a binary $0$-$1$ outcome,  the resulting probabilities have exponential form \@ref(eq:expo)  given by

@@

P_{\ak{\thetaN}}(\boldsymbol x) \propto
 \exp\left[\ak{\thetaN} \sum_{i=1}^N x_i\right], \qquad \boldsymbol x=(x_1,\ldots,x_N) \in\{0,1\}^N, 

@@

 with sufficient statistic $\boldsymbol g_N(\boldsymbol x)\equiv \sum_{i=1}^N  x_i$ and "log odds ratio" parameter $\ak{\thetaN} \equiv  \log[ P_{\ak{\thetaN}}(X_i=1)/P_{\ak{\thetaN}}(X_i=0) ] \in \mathbb{R}$. More generally, supposing $\boldsymbol X =(X_1,\ldots,X_N)$ represent $N$ independent trials, each assuming an outcome $\{1,\ldots,k\}$ among $k$ possibilities (e.g., a die roll), a multinomial distribution is given by

@@

P_{\ak{\thetaN}}(\boldsymbol x) \propto  \exp\left[  \ak{\thetaN}^T g_N(\boldsymbol x)   \right] =
\exp\left[ \sum_{j=1}^k {\theta_{j,\ak{\thetaidx}}} \sum_{i=1}^N \mathbb{I}(x_i=j) \right], \qquad \boldsymbol x  \in\{1,\ldots,k\}^N, 

@@

with sufficient statistic $\boldsymbol g_N(\boldsymbol x)$ involving a count $\sum_{i=1}^N \mathbb{I}(x_i=j)$ for each outcome $j \in \{1,\ldots,k\}$, where $\mathbb{I}(\cdot)$ denotes the indicator function, and parameters $\ak{\thetaN}=(\theta_{1,\ak{\thetaidx}},\ldots,\theta_{k,\ak{\thetaidx}})\in\mathbb{R}^k$ defining log-probability ratios $\theta_{i,\ak{\thetaidx}}-\theta_{j,\ak{\thetaidx}} =\log [P_{\ak{\thetaN}}(X_1=i)/P_{\ak{\thetaN}}(X_1=j)]$. In addition to such standard models for discrete independent data, exponential models of FOES type commonly arise with dependent spatial data [@besag1974spatial] and network/relational data [@wasserman1994social; @handcock2003assessing]. For a random graph  or network  with, say, $n$ nodes,  consider $N={n \choose 2}$ random edges where the $i$th edge is associated with a pair of nodes $s_i \equiv \{v_i,u_i\}$  and a binary variable $X_i\in\{0,1\}$ indicating   presence/absence of an edge among the node pair $s_i$, $i=1,\ldots,N$. Here the length $N$ of the edge variable sequence $\boldsymbol X = (X_1,\ldots,X_N)$ increases as a function of node number $n$ and corresponding exponential models often incorporate graph topographical features derived from $\boldsymbol X$. As an example, consider a graph model of exponential/FOES form prescribed by

@@

P_{\ak{\thetaN}}(\boldsymbol x) \propto
 \exp\left[\sum_{j=1}^3 \theta_{j,\ak{\thetaidx}} g_{j,N}(\boldsymbol x)\right], \quad\qquad \boldsymbol x=(x_1,\ldots,x_N)  \in\{0,1\}^N,

@@

involving the numbers of edges, 2-stars and triangles among an outcome $\boldsymbol x$ given by $g_{1,N}(\boldsymbol x)$, $g_{2,N}(\boldsymbol x)$ and $g_{3,N}(\boldsymbol x)$, respectively, along with $k=3$ real parameters $\ak{\thetaN} \equiv (\theta_{1,\ak{\thetaidx}},\theta_{2,\ak{\thetaidx}},\theta_{3,\ak{\thetaidx}})$. For this network model \@ref(eq:mod2) in particular, as well as for more general models of form \@ref(eq:expo),  @schweinberger2011instability considered instability in such exponential models with sequences of fixed parameters $\ak{\thetaN} = (\theta_1,\ldots,\theta_k)\in\mathbb{R}^k$, $N \geq 1$, of fixed dimension $k$.

For model sequences $P_{\ak{\thetaN}}(\boldsymbol x)\equiv p_{N,\ak{\thetaN}}(\boldsymbol x)$ of the exponential type \@ref(eq:expo), such as those in \@ref(eq:mod1)-\@ref(eq:mod2), note that the dimension $k$ of the parameter $\ak{\thetaN}\in\Theta \subset \mathbb{R}^k$ necessarily remains the same for all sample sizes $N \geq 1$ as the form of the natural parameter function $\eta(\cdot)$ in \@ref(eq:expo) and the number of sufficient statistics $\boldsymbol g_{N}(\boldsymbol x)$ do not depend on $N$. Consequently, $\ak{\thetaN}$ lies in a parameter space of fixed Euclidean dimension $k$. However, this aspect need not be true for other types of FOES models considered in Sections \@ref(rbm) - \@ref(deep-learning), where instead the numbers of parameters and sufficient statistics commonly increase with the sample size $N$.

@@

A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, with binary variables being most common [cf. @smolensky1986information]. A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with conditional independence within each layer. Let $\boldsymbol X = (X_1,\ldots,X_N)$ denote the $N$ random variables for visibles with support $\mathcal{X}^N$ and $\boldsymbol H = (H_1,\ldots,H_{N_\mathcal{H}})$ denote the $N_\mathcal{H}$ random variables for hiddens with support $\mathcal{X}^{N_\mathcal{H}}$ where $\mathcal{X} = \{-1,1\}$. For \ak{specified} parameters $\ak{\thetaN}^{\mathcal{H}} \in \mathbb{R}^{N_\mathcal{H}}$, $\ak{\thetaN}^{\mathcal{V}}\in \mathbb{R}^N$, and $\ak{\thetaN}^{\mathcal{HV}}$ as a real matrix with dimension $N_\mathcal{H} \times N$, the RBM model for $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)$  has the joint probability mass function

@@

\tilde{P}_{\ak{\thetaN}} (\tilde{\boldsymbol x}) = \exp\left[ (\ak{\thetaN}^{\mathcal{H}})^T \boldsymbol h + \boldsymbol (\ak{\thetaN}^{\mathcal{V}})^T \boldsymbol x + \boldsymbol h^T  \ak{\thetaN}^{\mathcal{HV}} \boldsymbol x - \psi(\ak{\thetaN})\right], \quad \tilde{\boldsymbol x} = (\boldsymbol x, \boldsymbol h) \in \{\pm 1\}^{N+N_\mathcal{H}}

@@

\psi(\ak{\thetaN}) = \log \sum_{\tilde{\boldsymbol x} \in \{\pm 1\}^{N+N_H} } \exp\left[ (\ak{\thetaN}^{\mathcal{H}})^T \boldsymbol h + \boldsymbol (\ak{\thetaN}^{\mathcal{V}})^T \boldsymbol x + \boldsymbol h^T  \ak{\thetaN}^{\mathcal{HV}} \boldsymbol x\right].

@@

Let $\ak{\thetaN} = (\ak{\thetaN}^{\mathcal{H}}, \ak{\thetaN}^{\mathcal{V}}, \ak{\thetaN}^{\mathcal{HV}} ) \in \Theta_\ak{\thetaidx} \equiv \mathbb{R}^{q(N)}$, with $q(N) = N + N_\mathcal{H} + N*N_\mathcal{H}$, denote \ak{a given} parameter vector for the RBM, as indexed by the number $N$ of visible random variables (which may differ from the actual lengths of these parameter vectors). The probability mass function for the visible variables $\boldsymbol X = (X_1, \dots, X_N)$ follows from marginalizing the joint specification to yield

@@

P_{\ak{\thetaN}} (\boldsymbol x) = \sum\limits_{\boldsymbol h \in \{\pm 1\}^{N_{\mathcal{H}}}} \tilde{P}_{\ak{\thetaN}} (\boldsymbol x, \boldsymbol h), \qquad \boldsymbol x \in \{\pm 1\}^{N}\equiv \mathcal{X}^N.

@@

Here the baseline model \@ref(eq:RBM1) for hidden/visible variables is a linear exponential one in sufficient statistics $(\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)$ using $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)$ from   \@ref(eq:RBM1), but the form differs from the previous exponential models in \@ref(eq:expo) in that the lengths of parameters $\ak{\thetaN}$ and statistics  $(\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)$ increase to incorporate more visible variables. That is, in contrast to \@ref(eq:expo), the natural parameter function involved in the RBM model \@ref(eq:RBM1), as the identity mapping of the parameters $\ak{\thetaN}\in\mathbb{R}^{q(N)}$, naturally grows in  dimension  $q(N)\to \infty$ to accommodate visible variables $X_1, \dots, X_N$ of increasing sample size $N\to \infty$. Additionally, one may further arbitrarily choose the number $N_\mathcal{H}$ of hidden variables $\boldsymbol H$ in the joint RBM model \@ref(eq:RBM1) to define a marginal model \@ref(eq:RBM2) for the $N$ visible variables $\boldsymbol X$, and the number $N_\mathcal{H}$ of hiddens may also potentially increase with $N$. Because $|\mathcal{X}| = 2$ and $P_{\ak{\thetaN}}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, the RBM  specification \@ref(eq:RBM2) for visibles $\boldsymbol$ corresponds to a FOES model, while the joint distribution \@ref(eq:RBM1) for $(\boldsymbol X, \boldsymbol H)$ is also a FOES model. As this example also indicates, any model formed by marginalizing a base FOES model class, such as the RBM joint specification \@ref(eq:RBM1), is again a FOES model.

@@

Consider two models with "deep architecture" that contain multiple hidden (or latent) layers in addition to a visible layer of data, namely a deep Boltzmann machine [@salakhutdinov2009deep] and a deep belief network [@hinton2006fast]. Let $M$ denote the number of hidden layers included in the model and let $N_{(H,1)}, \dots, N_{(H,M)}$ denote the numbers of hidden variables within each hidden layer. Then the random vector $\tilde{\boldsymbol X} = \{H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}, \dots, H^{(M)}_1, \dots, H^{(M)}_{N_{(H,M)}}, \boldsymbol X\}$ collects both the hidden variables $\{ H_{i}^{(j)} : i=1,\ldots, N_{(H,j)}, j=1,\ldots,M\}$ and visible variables $\boldsymbol X =(X_1,\ldots,X_N)$ in a deep probabilistic model. Each variable outcome will again lie in $\mathcal{X} = \{-1,1\}$.

**Deep Boltzmann machine (DBM).** The DBM class of models maintains conditional independence within all layers in the model by stacking RBM models and only allowing conditional dependence between neighboring layers. The joint probability mass function for a DBM is 
$$
\tilde{P}_{\ak{\thetaN}} ( \tilde{\boldsymbol x} ) = \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)} - \psi(\ak{\thetaN}) \right],
$$ 
for $\tilde{\boldsymbol x} = (\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}, \boldsymbol x) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}$ where 
$$
\psi(\ak{\thetaN}) = \log \sum\limits_{\tilde{\boldsymbol x} \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)} +N}} \exp\left[ \sum\limits_{i = 1}^M\boldsymbol \alpha^{(i)T} \boldsymbol h^{(i)} + \boldsymbol \beta^T \boldsymbol x + \boldsymbol h^{(1)T} \Gamma^{(0)} \boldsymbol x + \sum\limits_{i = 1}^{M - 1} \boldsymbol h^{(i)T} \Gamma^{(i)} \boldsymbol h^{(i + 1)}\right],
$$ 
is the normalizing function for $\ak{\thetaN} = (\boldsymbol \alpha^{(1)}, \dots, \boldsymbol \alpha^{(M)}, \boldsymbol \beta,\Gamma^{(0)}, \dots, \Gamma^{(M - 1)}) \in \Theta_\ak{\thetaidx} \subset \mathbb{R}^{q(N)}$, consisting of model parameters $\boldsymbol \beta \in \mathbb{R}^N$, $\boldsymbol \alpha^{(i)} \in \mathbb{R}^{N_{(H,i)}}$, $i = 1, \dots, M$, along with a matrix $\Gamma^{(0)}$ of dimension $N_{(H,1)} \times N$, and matrices $\Gamma^{(i)}$ of dimension
$N_{(H,i)} \times N_{(H,i+1)}$ for $i = 1, \dots, M-1$. The combined parameter vector $\ak{\thetaN}$ has total length $q(N)= N_{(H,1)}+\cdots N_{(H,M)} + N + N_{(H,1)}*N+N_{H,2}*H_{(H,1)}+\cdots +N_{(H,M)}*H_{(H,M)-1}$. The probability mass function for the visible random variables $X_1, \dots, X_N$ follows from this joint specification as 
$$
P_{\ak{\thetaN}} (\boldsymbol x) = \sum\limits_{(\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}) \in \mathcal{X}^{N_{(H,1)} + \cdots + N_{(H,M)}}} \tilde{P}_{\ak{\thetaN}} (\boldsymbol h^{(1)}, \dots, \boldsymbol h^{(M)}, \boldsymbol x) , \qquad \boldsymbol x \in \mathcal{X}^N.
$$ 
Again like the RBM case, the DBM model specification is an example of a FOES model.

**Deep belief network (DBN).** A DBN resembles a DBM in that there are multiple layers of latent random variables stacked in a deep architecture with no conditional dependence between layers. The difference between the DBM and DBN models is that all but the last stacked layer \ak{(layer $0$, which contains the visible random variables $X_1, \dots, X_N$)} in a DBN are Bayesian networks [see @pearl985bayesian], rather than RBMs. A Bayesian network is a probabilistic graphical model that defines conditional dependence to be directed, rather than undirected (as with the RBM).  Thus for visibles $X_1, \dots, X_N$ with support $\mathcal{X}^N, \mid \mathcal{X} \mid < \infty$, a DBN is also a FOES model with $q(N)$ the length of parameter vector is dependent on the dimension of the visibles because $P_{\ak{\thetaN}}(\boldsymbol x)>0$ for all $\boldsymbol x \in\mathcal{X}^N$. Commonly, as in logistic belief nets [@neal1992connectionist], a "weight" parameter is placed on each interaction between visibles, $X_1, \dots, X_N$, and the first layer of latent variables, $H^{(1)}_1, \dots, H^{(1)}_{N_{(H,1)}}$, satisfying the definition of a FOES model.
